---
title: "Sampling from a Bivariate Normal Distribution"
author: "Nikolai German"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We demonstrate how to transform a bivariate standard normal random variable $Z \sim \mathcal{N}_2(\boldsymbol{0}, I_2)$ into a general bivariate normal random variable $X \sim \mathcal{N}_2(\boldsymbol{\mu}, \Sigma)$ using the Cholesky decomposition.

## Mathematical Foundation

The transformation is based on:
$$X = \boldsymbol{\mu} + ZR$$
where $R$ is the Cholesky factor of the covariance matrix $\Sigma = R^T R$.

# Generating Standard Bivariate Normal Samples

First we construct $Z$ by sampling two independent standard normal variables:

```{r}
Z <- cbind(Z_1 = rnorm(100),
            Z_2 = rnorm(100))
```

## Verifying Standard Properties

We check that our samples have the expected properties of standard bivariate normal variables:

```{r}
# means
round(colMeans(Z), 2)
# variance
round(apply(Z, MARGIN = 2, var), 2)
# covariance
round(cov(Z[,1], Z[,2]), 2)
```

The results confirm approximately zero means, unit variances, and zero covariance.

# Setting Target Parameters

Next, we define the desired mean vector $\boldsymbol{\mu}$ and covariance matrix $\Sigma$:

```{r}
Sigma <- matrix(c(5, 2, 2, 1), ncol = 2)
mu <- c(-1, 1)
```

# Cholesky Decomposition

We compute the Cholesky factor. Note that R's `chol()` function returns an upper triangular matrix:

```{r}
(R <- chol(Sigma))
```

This gives us $\Sigma = R^T R$ where $R$ is upper triangular.

# Transformation

We apply the transformation $X = \boldsymbol{\mu} + ZR$:

```{r}
X <- t(mu + t(Z %*% R))
colnames(X) <- c("X_1", "X_2")
```

## Verifying Transformed Properties

We check that the transformed samples have the desired properties:

```{r}
round(colMeans(X), 2)
round(apply(X, MARGIN = 2, var), 2)
round(cov(X[,1], X[,2]), 2)
```

The results show that our transformation successfully produces samples with the target mean vector and covariance structure.

# Visualization

Finally, we visualize both the original standard bivariate normal samples and the transformed samples:

```{r}
library(ggplot2)
library(patchwork)

p_01 <- ggplot() +
  geom_density2d(data = as.data.frame(Z), aes(Z_1, Z_2)) +
  coord_fixed(ratio = 1, xlim = c(-6,6), ylim = c(-4,4)) +
  labs(x = latex2exp::TeX("$Z_1$"),
       y = latex2exp::TeX("$Z_2$"),
       title = "Z: before transformation") +
  theme_light()

p_02 <- ggplot() +
  geom_density_2d(data = as.data.frame(X), aes(X_1, X_2)) +
  coord_fixed(ratio = 1, xlim = c(-6,6), ylim = c(-4,4)) +
  labs(x = latex2exp::TeX("$X_1$"),
       y = latex2exp::TeX("$X_2$"),
       title = "X: after transformation") +
  theme_light()

(p <- (p_01 | p_02))
```

The visualization clearly shows how the Cholesky transformation changes both the location (mean) and the shape/orientation (covariance structure) of the bivariate normal distribution.

```{r include=FALSE}
ggsave("../plots/bivariate.png", p, device = "png", width = 12, height = 5)
```