---
title: "Sampling from a bivariate normal distribution"
author: "Nikolai German"
date: "`r Sys.Date()`"
output: html_document
---
We want to show how to transform a bivariate standard normal distributed random variable $Z \sim \mathcal{N}_2(\boldsymbol{0}, I_2)$ into a normal distributed random variable $X$ with expectation $\boldsymbol{\mu}$ and variance structure $\Sigma$.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we construct $Z$ by sampling two equal length vectors of i.i.d. standard normal samples.

```{r}
Z <- cbind(Z_1 = rnorm(100),
            Z_2 = rnorm(100))
```

Checking the properties of the samples:

```{r}
# means
round(colMeans(Z), 2)
# variance
round(apply(Z, MARGIN = 2, var), 2)
# covariance
round(cov(Z[,1], Z[,2]), 2)
```

Next, we initialize a vector of means $\boldsymbol{\mu}$ and a covariance matrix $\Sigma$.

```{r}
Sigma <- matrix(c(5, 2, 2, 1), ncol = 2)
mu <- c(-1, 1)
```

The implementation of Cholesky decomposition `chol()` returns an upper right triangular matrix, rather than a lower left:

$$
\Sigma = R^\top R
$$

```{r}
(R <- chol(Sigma))
```

Subsequently, we compute $X = \boldsymbol{\mu} + Z \cdot R$

```{r}
X <- t(mu + t(Z %*% R))
colnames(X) <- c("X_1", "X_2")
```

We have a look at the properties of $X$:

```{r}
round(colMeans(X), 2)
round(apply(X, MARGIN = 2, var), 2)
round(cov(X[,1], X[,2]), 2)
```

Lastly, we visualise $Z$ and $X$

```{r}
library(ggplot2)
library(patchwork)

p_01 <- ggplot() +
  geom_density2d(data = as.data.frame(Z), aes(Z_1, Z_2)) +
  coord_fixed(ratio = 1, xlim = c(-6,6), ylim = c(-4,4)) +
  labs(x = latex2exp::TeX("$Z_1$"),
       y = latex2exp::TeX("$Z_2$"),
       title = "Z: before transformation") +
  theme_light()

p_02 <- ggplot() +
  geom_density_2d(data = as.data.frame(X), aes(X_1, X_2)) +
  coord_fixed(ratio = 1, xlim = c(-6,6), ylim = c(-4,4)) +
  labs(x = latex2exp::TeX("$X_1$"),
       y = latex2exp::TeX("$X_2$"),
       title = "X: after transformation") +
  theme_light()

(p <- (p_01 | p_02))
```


```{r include=FALSE}
ggsave("../plots/bivariate.png", p, device = "png", width = 12, height = 5)
```

